Solving Collective Spin Hamiltonians with Decoherence

### 2018-08-27 soc_squeezing/cumulant_expansions.tex
moved initial conditions calculations to an appendix

The equations of motion for $S_\mu$, $S_\mu S_z$, and $S_\mu S_\nu$
(i.e.~where $\mu,\nu\in\set{+1,-1}$) under the OAT Hamiltonian
$H_{\t{OAT}}=\chi S_z^2$ are determined by
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_-^m S_z^n}
  = i\chi \sp{S_z^2, S_+^\ell S_-^m S_z^n}_-
  + \gamma \D_-\p{S_+^\ell S_-^m S_z^n},
\end{align}
where we can use identity \eqref{eq:comm_z_mu} to simplify
\begin{align}
  \sp{S_z^2, S_+^\ell S_-^m S_z^n}_-
  = \sp{S_z^2, S_+^\ell S_-^m}_- S_z^n
  = S_+^\ell S_-^m \p{\sp{\ell-m}^2 + 2\sp{\ell-m} S_z} S_z^n,
  \label{eq:S_z^2_comm}
\end{align}
and the effect of decoherence through $\D_-\p{S_+^\ell S_-^m S_z^n}$ is
worked out in Appendix \ref{sec:decoherence}.  In order to compute
squeezing of the OAT Hamiltonian in the presence of single-spin decay,
we therefore need to solve the systems of equations defined by
\begin{align}
  \f{d}{dt} \p{S_\mu S_z^n}
  = i\chi S_\mu \p{1 + 2\mu S_z} S_z^n
  + \gamma S_\mu \sp{\p{S - \delta_{\mu,-1} + S_z} \sum_{k=0}^{n-1}
    \p{-1}^{n-k} { n \choose k } S_z^k - \f12 S_z^n}
\end{align}
for $\mu\in\set{+1,-1}$, and
\begin{align}
  \f{d}{dt} \p{S_+^2 S_z^n}
  = 4i\chi S_\mu S_\nu \p{1 + S_z} S_z^n
  + \gamma S_\mu S_\nu \sp{\p{S + S_z} \sum_{k=0}^{n-1}
    \p{-1}^{n-k} { n \choose k } S_z^k - S_z^n},
\end{align}
while $d/dt\p{S_\pm S_\mp}=0$, so
\begin{align}
  \bk{S_\pm S_\mp} = \bk{S_\pm S_\mp}_{t=0} e^{-\gamma t}.
\end{align}
The initial conditions for these systems of equations can be
determined from the expansions
\begin{align}
  S_z^n = \sum_{m=-S}^S m^n \op{m},
  &&
  S_\mu = \sum_{m=-S}^S g_\mu\p{m} \op{m+\mu}{m},
\end{align}
with
\begin{align}
  g_\mu\p{S,m} \equiv \sqrt{\p{S-\mu m}\p{S+\mu m+1}},
  \label{eq:transverse_elem}
\end{align}
and the initial state
\begin{align}
  \ket{X} = 2^{-N/2} \sum_{m=-S}^S { N \choose S+m }^{1/2} \ket{m},
\end{align}
from which we can determine
\begin{align}
  \bk{S_\mu S_z^n}_{t=0}
  = 2^{-N} \sum_{m=-S}^S m^n { N \choose S+m } \abs{m - \mu S},
\end{align}
and
\begin{align}
  \bk{S_\mu S_\nu S_z^n}_{t=0}
  = 2^{-N} \sum_{m=-S+\delta_{\mu,-1}+\delta_{\nu,-1}}
  ^{S-\delta_{\mu,1}-\delta_{\nu,1}}
  m^n \sp{{ N \choose S+m+\mu+\nu } { N \choose S+m }}^{1/2}
  g_\mu\p{S,m+\nu} g_\nu\p{S,m}.
\end{align}

@@

All collective spin Hamiltonians can be written in the form
\begin{align}
  H = \sum_{p,q,r} h_{pqr} S_+^p S_-^q S_z^r.
  \label{eq:H_general}
\end{align}
The Hamiltonian for one-axis twisting with a transverse-field, for
example, is
\begin{align}
  H_{\t{TVF}}
  = \chi S_z^2 + \Omega S_x
  = \chi S_z^2 + \f{\Omega}{2} S_+ + \f{\Omega}{2} S_-,
\end{align}
while the two-axis twisting Hamiltonians about $\p{z,y}$ and $\p{y,x}$
are
\begin{align}
  H_{\t{TAT}}^{z,y}
  = \chi \p{S_z^2 - S_y^2}
  = \chi \p{S_z^2 + \f14 S_+^2 + \f14 S_-^2 - \f12 S_+ S_- + \f12 S_z},
\end{align}
and
\begin{align}
  H_{\t{TAT}}^{y,x}
  = \chi\p{S_y^2 - S_x^2}
  = -\f{\chi}{2} \p{S_+^2 + S_-^2}.
\end{align}
The evolution of a general operator $S_+^\ell S_-^m S_z^n$ under a
Hamiltonian of the form in \eqref{eq:H_general} is given by
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_-^m S_z^n}
  = i\sum_{p,q,r} h_{pqr} \sp{S_+^p S_-^q S_z^r, S_+^\ell S_-^m S_z^n}_-
  + \gamma \D_-\p{S_+^\ell S_-^m S_z^n}
  = \sum_{p,q,r} T^{\ell m n}_{pqr} S_+^p S_-^q S_z^r,
\end{align}
where the commutator is given in Appendix \ref{sec:comm_general}, and
the decoherence operator in Appendix \ref{sec:decoherence}.  We will
be considering evolution of the initial states
$\ket{X}:S_x\ket{X}=S\ket{X}$ and $\ket{-Z}:S_z\ket{-Z}=-S\ket{-Z}$,
for which initial conditions are given by
\begin{multline}
  \bk{X|S_+^\ell S_-^m S_z^n|X}
  =  2^{-N} \sum_{k=-S+m}^{S-\ell} k^n
  \sp{{ N \choose S+k+\ell-m } { N \choose S+k }}^{1/2} \\
  \times \prod_{p=0}^{\ell-1} g_+\p{S,k-m+p}
  \prod_{q=0}^{m-1} g_-\p{S,k-q}
\end{multline}
and
\begin{multline}
  \bk{-Z|S_+^\ell S_-^m S_z^n|-Z}
  = \p{-S}^n 2^{-N} \sum_{k=-S+m}^{S-\ell}
  \sp{{ N \choose S+k+\ell-m } { N \choose S+k }}^{1/2} \\
  \times \prod_{p=0}^{\ell-1} g_+\p{S,k-m+p}
  \prod_{q=0}^{m-1} g_-\p{S,k-q}.
\end{multline}
Equipped with these initial conditions, we can expand
\begin{align}
  \bk{S_+^\ell S_-^m S_z^n}
  = \sum_{k\ge0} \f{t^k}{k!}
  \bk{\f{d^k}{dt^k}\p{S_+^\ell S_-^m S_z^n}}_{t=0}
  \approx \sum_{k=0}^M \f{t^k}{k!}
  \bk{\f{d^k}{dt^k}\p{S_+^\ell S_-^m S_z^n}}_{t=0},
\end{align}
where $M$ is chosen such that
\begin{align}
  \f{t^M}{M!} \bk{\f{d^M}{dt^M}\p{S_+^\ell S_-^m S_z^n}}_{t=0}
  \ll S^{\ell+m+n}.
\end{align}
% comment on explosion of operators

@@

>>>>

The equations of motion for $S_\mu$, $S_\mu S_z$, and $S_\mu S_\nu$
(i.e.~where $\mu,\nu\in\set{+1,-1}$) under the OAT Hamiltonian
$H_{\t{OAT}}=\chi S_z^2$ are determined by
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_-^m S_z^n}
  = i\chi \sp{S_z^2, S_+^\ell S_-^m S_z^n}_-
  + \gamma \D_-\p{S_+^\ell S_-^m S_z^n},
\end{align}
where we can use identity \eqref{eq:comm_z_mu} to simplify
\begin{align}
  \sp{S_z^2, S_+^\ell S_-^m S_z^n}_-
  = \sp{S_z^2, S_+^\ell S_-^m}_- S_z^n
  = S_+^\ell S_-^m \p{\sp{\ell-m}^2 + 2\sp{\ell-m} S_z} S_z^n,
  \label{eq:S_z^2_comm}
\end{align}
and the effect of decoherence through $\D_-\p{S_+^\ell S_-^m S_z^n}$ is
worked out in Appendix \ref{sec:decoherence}.  In order to compute
squeezing of the OAT Hamiltonian in the presence of single-spin decay,
we therefore need to solve the systems of equations defined by
\begin{align}
  \f{d}{dt} \p{S_\mu S_z^n}
  = i\chi S_\mu \p{1 + 2\mu S_z} S_z^n
  + \gamma S_\mu \sp{\p{S - \delta_{\mu,-1} + S_z} \sum_{k=0}^{n-1}
    \p{-1}^{n-k} { n \choose k } S_z^k - \f12 S_z^n}
  \label{eq:diff_mu_OAT}
\end{align}
for $\mu\in\set{+1,-1}$, and
\begin{align}
  \f{d}{dt} \p{S_+^2 S_z^n}
  = 4i\chi S_\mu S_\nu \p{1 + S_z} S_z^n
  + \gamma S_\mu S_\nu \sp{\p{S + S_z} \sum_{k=0}^{n-1}
    \p{-1}^{n-k} { n \choose k } S_z^k - S_z^n},
  \label{eq:diff_pp_OAT}
\end{align}
while $d/dt\p{S_\pm S_\mp}=0$, so
\begin{align}
  \bk{S_\pm S_\mp} = \bk{S_\pm S_\mp}_{t=0} e^{-\gamma t}.
\end{align}
The initial conditions for \eqref{eq:diff_mu_OAT} and
\eqref{eq:diff_pp_OAT} are provided in Appendix
\ref{sec:initial_conditions}.

@@

All collective spin Hamiltonians can be written in the form
\begin{align}
  H = \sum_{p,q,r} h_{pqr} S_+^p S_-^q S_z^r.
  \label{eq:H_general}
\end{align}
The Hamiltonian for one-axis twisting with a transverse-field, for
example, is
\begin{align}
  H_{\t{TVF}}
  = \chi S_z^2 + \Omega S_x
  = \chi S_z^2 + \f{\Omega}{2} S_+ + \f{\Omega}{2} S_-,
\end{align}
while the two-axis twisting Hamiltonians about $\p{z,y}$ and $\p{y,x}$
are
\begin{align}
  H_{\t{TAT}}^{z,y}
  = \chi \p{S_z^2 - S_y^2}
  = \chi \p{S_z^2 + \f14 S_+^2 + \f14 S_-^2 - \f12 S_+ S_- + \f12 S_z},
\end{align}
and
\begin{align}
  H_{\t{TAT}}^{y,x}
  = \chi\p{S_y^2 - S_x^2}
  = -\f{\chi}{2} \p{S_+^2 + S_-^2}.
\end{align}
The evolution of a general operator $S_+^\ell S_-^m S_z^n$ under a
Hamiltonian of the form in \eqref{eq:H_general} is given by
\begin{align}
  \f{d}{dt} \p{S_+^\ell S_-^m S_z^n}
  = i\sum_{p,q,r} h_{pqr} \sp{S_+^p S_-^q S_z^r, S_+^\ell S_-^m S_z^n}_-
  + \gamma \D_-\p{S_+^\ell S_-^m S_z^n}
  = \sum_{p,q,r} T^{\ell m n}_{pqr} S_+^p S_-^q S_z^r,
\end{align}
where the commutator is given in Appendix \ref{sec:comm_general}, and
the decoherence operator in Appendix \ref{sec:decoherence}.  We can
then expand
\begin{align}
  \bk{S_+^\ell S_-^m S_z^n}
  = \sum_{k\ge0} \f{t^k}{k!}
  \bk{\f{d^k}{dt^k}\p{S_+^\ell S_-^m S_z^n}}_{t=0}
  \approx \sum_{k=0}^M \f{t^k}{k!}
  \bk{\f{d^k}{dt^k}\p{S_+^\ell S_-^m S_z^n}}_{t=0},
\end{align}
where $M$ is chosen such that
\begin{align}
  \f{t^M}{M!} \bk{\f{d^M}{dt^M}\p{S_+^\ell S_-^m S_z^n}}_{t=0}
  \ll S^{\ell+m+n}.
\end{align}
Initial conditions, i.e.~expectation values at $t=0$, are provided in
Appendix \ref{sec:initial_conditions}.
% comment on explosion of operators

@@

\section{Initial conditions}
\label{sec:initial_conditions}

Initial conditions for all collective spin operators can be determined
from the expansions
\begin{align}
  S_z^n = \sum_{m=-S}^S m^n \op{m},
  &&
  S_\mu = \sum_{m=-S}^S g_\mu\p{m} \op{m+\mu}{m},
\end{align}
with
\begin{align}
  g_\mu\p{S,m} \equiv \sqrt{\p{S-\mu m}\p{S+\mu m+1}},
\end{align}
and the initial state
\begin{align}
  \ket{X} \equiv 2^{-N/2} \sum_{m=-S}^S { N \choose S+m }^{1/2} \ket{m},
\end{align}
using which we can compute
\begin{multline}
  \bk{X|S_+^\ell S_-^m S_z^n|X}
  =  2^{-N} \sum_{k=-S}^{S-\max\set{\ell,m}} \p{k+m}^n
  \sp{{ N \choose S+k+\ell } { N \choose S+k+m }}^{1/2} \\
  \times \prod_{p=0}^{\ell-1} g_+\p{S,k+p}
  \prod_{q=0}^{m-1} g_+\p{S,k-q}.
\end{multline}
